Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-11-13 main.tex
Added a note on the mixed effects model and the ICC being very low for reviewer #2.

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt.
These two fall criteria are not strictly the same and the latter is more likely
to happen at higher speeds.
We informed the participants beforehand that we would stop the treadmill in
these situations.
These were the two actions which necessitated stopping the treadmill belt and
the simulate real fall modes.
Placing a foot down on the fall side of the bicycle is a natural reaction when
steering or other body movements are not sufficient to keep the bicycle
upright, as shown in \citep{Gildea2024a}.
Exiting the width of the treadmill is akin to exiting the width of a curb-less
bicycle path and arises when the rider's control actions are not sufficient to
correct the heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \citet{Enders2007}, with clusters being an
individual participant. The clusters are chosen as all data from an individual
participant because we are only interested in the association between the state
of the balance assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation; however, this was not true. Cluster-mean centering
showed there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and one
for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis for each
set of data. Statistics were computed with the lme4 version
1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.

>>>>

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt.
These two fall criteria are not strictly the same and the latter is more likely
to happen at higher speeds.
We informed the participants beforehand that we would stop the treadmill in
these situations.
These were the two actions which necessitated stopping the treadmill belt and
the simulate real fall modes.
Placing a foot down on the fall side of the bicycle is a natural reaction when
steering or other body movements are not sufficient to keep the bicycle
upright, as shown in \citep{Gildea2024}.
Exiting the width of the treadmill is akin to exiting the width of a curb-less
bicycle path and arises when the rider's control actions are not sufficient to
correct the heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering with clusters being an individual participant, as recommended by
\citet{Enders2007}. The clusters are chosen as all data from an individual
participant because we are only interested in the association between the state
of the balance assist system and the outcome of the perturbation for that
participant, not how participants perform relative to each other. We expected
there to be variation between participants in how well they are able to resist
the perturbation; however, this was not true. If a random effect is included in
the model for each participant, the intraclass correlation (ICC) is less that
3\% for both the 6~\si{\kph} and 10~\si{\kph} datasets, so we utilize a simple
single-level logistic regression model without the random effect instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
for each set of data. Statistics were computed with the lme4 version
1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.
